---
title: "Script3_2conditions_FitnessComparison"
author: "Manon Morin"
date: "6/29/2020"
output: html_document
---
##**STEP 3: Comparison of fitness values and identification of interaction fitness **

## Script: Script3_2condidtions_FitnessComparison.Rmd
       
        
Compares gene fitness values of all conditions against a chosen reference condition. Appends a “Category” to each compared gene to indicate if fitness values are significantly different or not based on chosen statistical criteria.

<br>

### Run
#### Packages and functions 

```{r, include=TRUE}
rm(list=ls())

# Package and functions upload

library(ggplot2)
library(dplyr)
library(tidyr)
library(gridExtra)

source("Category_definition.R")
source("Comparison_test.R")
```
<br>

#### Data upload and parameters set up

```{r, include=TRUE, echo=TRUE}

# Step 1: Data import and parameters settings
  # Import the .Rdata file generated in Averaging_replicates.R
  # Set up org_locId
  # Set up your condition of reference (has to be one of your conditions)
  # Set up your alpha value for the T-test
  # Set up whether you want to performed correction for multiple comparison testing and screen on adjusted-pvalue

load(".RData")   # here you upload the .RData ouput of Script2_Averaging_replicates.R
Data_Fitness=Final_gene_Fitness

org_locId=""  # Again, you set up your organisms whether "Ecoli" if the locusId are in numeric form or "Pseudo" if the locusId are in character form
Condition1=""  #here you write the name of one of the 2 conditions you want to compare. Make sure it is the same name than previously used
alphaF=0.002    #here you choose any alpha you want for the Fisher test (Test for equal variance) You can choose 0.05 or 0.002. 
alphaT=0.05   #here you choose any alpha you want. Just be aware that it is where you can control the amount of false discovery you allow
multi=1    # here you decide if you want to correct for multiple comparison (method=fdr) multi = 0 ==> no correction; multi=1 correction


```
<br>

#### 2 conditions comparison against a chosen reference condition (Condition 1)

<span style="color:darkslateblue">
*In the following plots: "Sig" means that gene fitness values are significantly different for that gene between the compared conditions, "Not_Sig" means that fitness values are not significantly different, and "Not_tested" means that the comparison has not been performed for that gene due to unequal variances*
</span>

```{r,include=TRUE, echo=TRUE}

# Step 2: Run the comparison for all conditions against the reference one
  # Each conditions that is not the reference is going to be tested against the refence one after the other using a for loop embedding the comparison function


Table_all=data.frame() # creates a table to eventually store the data
List_plots=list(0,0,0,0,0,0,0) # creates a list to store comparison of plots for each comparison in the "for loop" . 
                               # needs to have as many spots in the list than you have comparisons

Conditions=as.vector(unique(Data_Fitness$Cdt))
Conditions=Conditions[Conditions!=Condition1]# generates a vector with all the conditions to compare against the reference one

for (i in 1:length(Conditions)){
  Cdt=Conditions[i]
  Condition2=paste(Cdt)  
  Test1=Comparison_test(Data_Fitness,Condition1, Condition2, alphaT, alphaF, org_locId, multi)   
  #You obtain a list
  #list[[1]]= the table containing all the genes tested + fitness + variance values in both conditions + Fcalc (Fisher value calculates for the test of variance + Tcalc (Tscore value for the student test))
  # and the category column indicates if the difference of fitness is significant (Sig), if it is not (Not_Sig) or if the student was not performed because of unequal variances (Not_tested)
  #list[[2]]=the scatter plot representing the data
  
  Table_all=rbind(Table_all, Test1[[1]])
  List_plots[[i]]= Test1[[2]]
}

#Visualize all the scatter plots
grid.arrange(List_plots[[1]],List_plots[[2]],List_plots[[3]],List_plots[[4]], nrow=2,ncol=2)

write.csv(Table_all, ".csv")
save(Table_all, List_plots, file=".RData")


```
